Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-10-17 manu-draft.qmd
more edits from rn comments

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes. These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida. Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point). Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st.

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package \[@Beck21\], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.

The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa. The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a). The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air–water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity. The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c). There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series (0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$). There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$. As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

# gasex nse
prmall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  filter(var != 'DO_mod') %>% 
  na.omit %>% 
  summarise(
    nse = round(median(nse), 2),
    .by = var
  ) %>% 
  tibble::deframe()

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). This result makes sense, given that the $b$ parameter was fixed at the actual value used to create the synthetic time series. Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`).

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were dervied from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the residual and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors justified in @sec-ebasemod (equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

Pairwise comparisons of the daily estimates of $NEM$, $P$, and $R$ returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square differences (RMSD) of the linear regression fit comparing the same flux type between methods. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series. @fig-apacmpfig and @tbl-apacmptab provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected. However, correlations were generally stronger (and RMSDs smaller) for the metabolic results based on the detided dissolved oxygen time series. The observed time series clearly produced a much larger range of values for results from both the EBASE and Odum methods.

@@

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over- and under-estimates of metabolic rates with more variation than those without tidal influence (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, more variable estimates are returned. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework. This has the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.

There were several considerations during the development of EBASE that makes the model appropriate for estuarine settings. As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. The most critical change was the formulation for gas exchange in EBASE using the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. EBASE also removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day; e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter (such as fetch and waves) as well as ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across sytems. For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized. However, an advantage of the Bayesian approach is the ability to evaluate different optimization periods and, therefore, the amount of data on which to estimate key parameters. Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.). The R package accommodates data gaps using linear interpolation for any required input used by EBASE. This allows the Bayesian routine to provide continuous estimates across the time series. The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations. The analyst can determine the appropriate length of time to use for excluding results based, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods in a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries. Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.

>>>>

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes. These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida. Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point). Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st [@akaike1969].

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.

The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa. The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a). The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity. The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c). There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$. As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

# all parameter nse
prmall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  na.omit %>% 
  summarise(
    nse = round(median(nse), 2),
    .by = var
  ) %>% 
  tibble::deframe()

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons. After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`. This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$. Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.

@@

#### Impact of noise on parameter recovery

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors justified in @sec-ebasemod (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

Pairwise comparisons of the daily estimates of $NEM$, $P$, and $R$ returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square differences (RMSD) of the linear regression fit comparing the same flux type between methods. @fig-apacmpfig and @tbl-apacmptab provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected. However, correlations were generally stronger (and RMSDs smaller) for the metabolic results based on the detided dissolved oxygen time series. The observed time series clearly produced a much larger range of values for results from both the EBASE and Odum methods.

@@

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over- and under-estimates of metabolic rates with more variation than those without tidal influence (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, more variable estimates are returned. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework. Such an approach would have the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.

As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former. The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE. EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization periods, whereas BASE was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across systems. For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized. However, an advantage of the Bayesian approach is the ability to evaluate different optimization periods and, therefore, the amount of data on which to estimate key parameters. Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.

The EBASE R package can also accommodate missing observations in the input data. This feature allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction and routine maintenance). The R package accommodates data gaps using linear interpolation for any required input used by EBASE, allowing the Bayesian routine to provide continuous estimates across the time series. The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations. The analyst can determine the appropriate length of time to use for excluding results, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods in a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to constrain the gas transfer velocity from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries. Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
